Accelerating Deep Neural Network Training on FPGAs Facilitating the Use of Variable Precision

### 2018-06-28 SWoPP2018 Paper/paper.tex
Fixed some spacing issues.

The recent explosion in artificial intelligence – in particular that of deep neural networks based on backpropagation – has triggered a storm in the introduction and creation of specialized compute devices. Commercial platforms such as Microsoft’s BrainWave~\cite{msbrainwave}, Google’s TPUs~\cite{googletpu}, and Fujitsu’s DLU~\cite{fujitsudlu} are all examples of specialized circuitry dedicated to low-power, high-performance Deep-Learning (DL) training and/or inference. Meanwhile, existing general-purpose manufacturers are empowering their architecture with custom floating-point units that trade precision for performance, such as Intel’s Knights-Mill~\cite{knm} and NVIDIA’s Volta-100~\cite{volta100}. It is clear that Artifical Intelligence and Deep-Learning will have a prioritized presence in modern architecture -- today and in the near future.

@@

However, exploring the space around IEEE-754 floating-point representations and its alternatives pose a significant engineering problem: hardware and software infrastructure is to tightly coupled to the IEEE-754 standard that changing any part of it incurs a high engineering overhead.
One alternative is to simulate alternative floating-point representations in software, for example using soft-float or MPFR libraries~\cite{softfloat}, but the resulting performance is often several magnitudes lower than hardware implementations, limiting the size of the study conducted. Furthermore, simulating alternative numerical formats in software are incapable of using any compiler optimizations, nor can they leverage the vector instruction often crucial to reach application performance in modern processors. Porting the full software infrastructure stack is possible (and inevitable), but is a non-trivial effort that requires changes to the compiler, standard libraries and (possibly) the Application-Binary-Interface (e.g. calling convention).

@@

The remainder of our paper is structure in the following way: Section~\ref{sec:relwork} position our work against other, similar efforts. Section~\ref{sec:framework} gives a background to machine learning and FPGAs, and describes our proposed architecture. Section~\ref{sec:method} overviews our experimental methodology, and is followed by our preliminary results in section~\ref{sec:result}. We conclude in section~\ref{sec:conclusion}.

\section{Related Work}
\label{sec:relwork}
Field-Programmable Gate-Arrays (FPGAs) have extensively been used to probe and explore various architectural design decision, including those of deep-learning. The large majority of existing work limit themselves to inference, primarily due to the simpler design layout (little intermediate data needs storing). Levering FPGAs have allowed researchers to decrease the number of bits allocate to the numerical representation, going as far as inferring popular networks such as  AlexNET~\cite{krizhevsky2012imagenet} using as few as binary~\cite{shimoda2017all,umuroglu2017finn} (1-bit) weights. Inferring networks that uses few bits to represents weights (called Quantized Neural Networks~\cite{courbariaux2016binarized}) requires subtle yet necessary changes to the training phase~\cite{binaryconnect,courbariaux2016binarized,zhou2016dorefa}. Lately, modern deep-learning framework are having support for reduced precision training, some driven by leading FPGA vendors such as Xilinx~\cite{xilinxml}.
Dicecco et al.~\cite{dicecco2017fpga} are are working on a framework similar to ours – training neural networks using FPGAs while modifying the numerical format. Their work primary focus on the IEEE-754 format and convolution layers where-as we aspire to include alternative format such as posits~\cite{posits}. Using posits in deep-learning was intended the focus of Langroudi et al.~\cite{langroudi2018deep}’s work; however, they only compared the performance against fixed-point and only \textit{stored} the weights as posit (the computation was still done using IEEE-754).

@@

\section{A FPGA-based CNN Framework}
\label{sec:framework}
\subsection{Field-Programmable Gate-Arrays}
%Small intro to FPGAs.
FPGAs can be used for a great variety of things, from accelerating computational algorithms to implementating a hardware design before taping out full chip.
They are devices that can represent reconfigurable hardware. 
The fabric they are made up of consists of vast arrays of switches and other routing hardware to connect all the functional elements, like multipliers, block RAM and look up tables, in any way possible.
This makes them very suitable for the research described in this paper.
It gives total freedom to design a hardware implementation to explore algorithms in a hardware setting.

@@

\section{Methodology}\label{sec:method}
\subsection{Experimental Platform}
\begin{table}[h]\label{tab:software-versions}
  \caption{Used software versions}
  \begin{tabular}{ll}
    \toprule
    \textbf{Software} & \textbf{Version} \\
    \midrule
    Python & v3.6.3 \\
    Keras & v2.2.0 \\
    TensorFlow & v1.8.0 \\
    Intel Quartus & v17.1.0 \\
    \bottomrule
  \end{tabular}
\end{table}
\begin{table}[h]\label{tab:test-hardware}
  \caption{Test hardware}
  \begin{tabular}{ll}
    \toprule
    \textbf{Hardware} & \textbf{Configuration} \\
    \midrule
    Intel i7-3930K & 3.20 GHz \\
    Arria 10 & Nallatech N510T \\
    TensorFlow & v1.8.0 \\
    Intel Quartus & v17.1.0 \\
    \bottomrule
  \end{tabular}
\end{table}
Compiler, version, flags, fpga boards, host, etc.
\subsection{Network Tested + Data-set}
The used example network is a simple multilayer perceptron.
The main dataset used was MNIST with a layer depth of 1 and the first layer output size of 128.
The data set was trained using the first \num{50000} images from the MNIST dataset.
The last \num{10000} images are used as the test dataset.
Training was done using a batch size of 128.

>>>>

The recent explosion in artificial intelligence – in particular that of deep neural networks based on backpropagation – has triggered a storm in the introduction and creation of specialized compute devices. Commercial platforms such as Microsoft’s BrainWave~\cite{msbrainwave}, Google’s TPUs~\cite{googletpu}, and Fujitsu’s DLU~\cite{fujitsudlu} are all examples of specialized circuitry dedicated to low-power, high-performance Deep-Learning (DL) training and/or inference. Meanwhile, existing general-purpose manufacturers are empowering their architecture with custom floating-point units that trade precision for performance, such as Intel’s Knights-Mill\`\cite{knm} and NVIDIA’s Volta-100\`\cite{volta100}. It is clear that Artifical Intelligence and Deep-Learning will have a prioritized presence in modern architecture -- today and in the near future.

@@

However, exploring the space around IEEE-754 floating-point representations and its alternatives pose a significant engineering problem: hardware and software infrastructure is to tightly coupled to the IEEE-754 standard that changing any part of it incurs a high engineering overhead.
One alternative is to simulate alternative floating-point representations in software, for example using soft-float or MPFR libraries~\cite{softfloat}, but the resulting performance is often several magnitudes lower than hardware implementations, limiting the size of the study conducted. Furthermore, simulating alternative numerical formats in software are incapable of using any compiler optimizations, nor can they leverage the vector instruction often crucial to reach application performance in modern processors. Porting the full software infrastructure stack is possible (and inevitable), but is a non-trivial effort that requires changes to the compiler, standard libraries and (possibly) the Application-Binary-Interface (e.g.\`calling convention).

@@

The remainder of our paper is structure in the following way: Section~\ref{sec:relwork} position our work against other, similar efforts. Section~\ref{sec:framework} gives a background to machine learning and FPGAs, and describes our proposed architecture. Section~\ref{sec:method} overviews our experimental methodology, and is followed by our preliminary results in Section\`\ref{sec:result}. We conclude in section~\ref{sec:conclusion}.

\section{Related Work}\label{sec:relwork}
Field-Programmable Gate-Arrays (FPGAs) have extensively been used to probe and explore various architectural design decision, including those of deep-learning. The large majority of existing work limit themselves to inference, primarily due to the simpler design layout (little intermediate data needs storing). Levering FPGAs have allowed researchers to decrease the number of bits allocate to the numerical representation, going as far as inferring popular networks such as  AlexNET~\cite{krizhevsky2012imagenet} using as few as binary~\cite{shimoda2017all,umuroglu2017finn} (1-bit) weights. Inferring networks that uses few bits to represents weights (called Quantized Neural Networks~\cite{courbariaux2016binarized}) requires subtle yet necessary changes to the training phase~\cite{binaryconnect,courbariaux2016binarized,zhou2016dorefa}. Lately, modern deep-learning framework are having support for reduced precision training, some driven by leading FPGA vendors such as Xilinx~\cite{xilinxml}.
Dicecco et al.~\cite{dicecco2017fpga} are are working on a framework similar to ours – training neural networks using FPGAs while modifying the numerical format. Their work primary focus on the IEEE-754 format and convolution layers where-as we aspire to include alternative format such as posits~\cite{posits}. Using posits in deep-learning was intended the focus of Langroudi et al.~\cite{langroudi2018deep}’s work; however, they only compared the performance against fixed-point and only \textit{stored} the weights as posit (the computation was still done using IEEE-754).

@@

\section{A FPGA-based CNN Framework}\label{sec:framework}
\subsection{Field-Programmable Gate-Arrays}
%Small intro to FPGAs.
FPGAs can be used for a great variety of things, from accelerating computational algorithms to implementating a hardware design before taping out full chip.
They are devices that can represent reconfigurable hardware. 
The fabric they are made up of consists of vast arrays of switches and other routing hardware to connect all the functional elements, like multipliers, block RAM and look up tables, in any way possible.
This makes them very suitable for the research described in this paper.
It gives total freedom to design a hardware implementation to explore algorithms in a hardware setting.

@@

\section{Methodology}\label{sec:method}
\subsection{Experimental Platform}
\begin{table}[h]\label{tab:software-versions}
  \caption{Used software versions}
  \begin{tabular}{ll}
    \toprule
    \textbf{Software} & \textbf{Version} \\
    \midrule
    Python & v3.6.3 \\
    Keras & v2.2.0 \\
    TensorFlow & v1.8.0 \\
    Intel Quartus & v17.1.0 \\
    \bottomrule
  \end{tabular}
\end{table}
\begin{table}[h]\label{tab:test-hardware}
  \caption{Test hardware}
  \begin{tabular}{ll}
    \toprule
    \textbf{Hardware} & \textbf{Configuration} \\
    \midrule
    Intel i7-3930K & 3.20 GHz \\
    Arria 10 & Nallatech N510T \\
    TensorFlow & v1.8.0 \\
    Intel Quartus & v17.1.0 \\
    \bottomrule
  \end{tabular}
\end{table}
Compiler, version, flags, fpga boards, host, etc.
\subsection{Network Tested + Data-set}
The used example network is a simple multilayer perceptron.
The main dataset used was MNIST with a layer depth of 1 and the first layer output size of 128.
The data set was trained using the first \num{50000} images from the MNIST dataset.
The last \num{10000} images are used as the test dataset.
Training was done using a batch size of 128.
The CPU code was implemented using the NumPy array.dot function.
